---
title: "BC AF Trade Tool Build"
author: "Leila Bautista and Brendan Dwyer"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# BC AF Trade Tool Build

This RMarkdown provides chunks of R code to load, filter and combine data from Open Government [website](https://search.open.canada.ca/opendata/?sort=metadata_modified+desc&search_text=cimt&page=1). The trade data is used by the Ministry of Agriculture and Food's SITS Unit to produce various deliverables that estimate BC and other provincial trade of agriculture and food commodities.

Contact [alstats\@gov.bc.ca](mailto:alstats@gov.bc.ca) for any questions. The following are key steps:

1.  Set-up and load library

2.  Loop through 2003-2023 to load and read data on:

    a\. domestic export\*

    b\. import

3.  Calculate trade balance

4.  Append all years into the time series data called [**TradeFull.csv**]{.underline}

\*A new chunk to load **c. Total export** could be written which would require binding with **a. Domestic export** to pivot wide by export type (i.e., total vs. domestic). Once pivoted wide, re-export is equal to total export less domestic export and can be calculated. After calculating re-export, optional to drop total export and pivot long to match **Trade_Update** and **Trade_Full** data frames' variables (i.e., columns) and types. Section **3. Calculate the trade balance** will need to be updated to allow for the distinction between domestic, re-export or total-export. Additional work on the **BC_AF_Trade_Tool_Update.Rmd** will also need to be undertaken.

# 1. Set-up & Library

As of January 2026, SITS cannot yet directly connect to a SharePoint URL to access the shared Unpublished Data library. The interim workaround is to sync that library to the user's Desktop and change the URL manually. Remember to replace "\\" with "/" and note that the timeout settings is 10 mins because the zip file is large and the default timeout of 1 min is insufficient to download and unzip the file.

```{r Set-up, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory

knitr::opts_knit$set(root.dir = "C:/Users/lebautis/Government of BC/SICI Data Library Subsite - Unpublished Data")

# sets the time-out option to 10 mins
options(timeout = 600)
```

```{r Library}

# Load library
library(pacman)
p_load(dplyr,
       data.table,
       tidyr,
       lubridate,
       tictoc,
       splitstackshape,
       stringr)
```

# 2. Loop from 2003-2023 to load and read data

As of January 2026, there are 2 chunks of code involved to load the update for each of:

-   a\. Domestic export

-   b\. Import

See the top of this .Rmd for more information on total and re-export integration.

## 2a. Domestic Export Historic

This chunk of code works off the user's desktop and the directory will need to be updated with their IDIR:

```         
"C:/Users/[replace with IDIR]/Desktop/"
```

The URL for the .zip file changes and needs updating. Since 2017, it has been formatted like:

```         
https://www150.statcan.gc.ca/n1/pub/##-###-x/#######/zip/CIMT-CICM_Dom_Exp_YYYY.zip
```

From 2003 to 2016, the format of the .zip file was:

Confirm that file_inside_zip is the correct one to pull from the zip_contents then update the references for [**out_dir**]{.underline} and [**file.path**]{.underline} with the latest year:

```         
C:/Users/[replace with IDIR]/Desktop/CIMT_CICM_YYYY/CIMT-CICM_Dom_Exp_YYYY
```

Until BC Gov admins grant access, SharePoint files cannot easily be pulled using the **M365** R package. The temporary workaround is to sync the Unpublished Data Library with each user's desktop which needs to be updated in the code:

```         
"C:/Users/[replace with IDIR]/Government of BC/SICI Data Library Subsite - Unpublished Data"
```

When all of the above references have been updated, following chunk will:

-   Download the zip file: load the specific file with export value data at the HS8-level into a full data frame called **DomExport_Update**.

-   Clean the data: rename columns and specifying types (e.g., text, numeric, etc.)

-   Filter the full data frame: load the specific list of HS8 codes from AAFC's CATSNET related to agriculture and food into a data framed called **Commodity_REF** and left join with the full data frame, **DomExport_Update,** to filter only for matching HS8 codes

-   Remember to replace "\\" with "/"

```{r Domestic Export Historic}

tic()

# Load the list of agriculture, seafood and food manufacturing HS codes 

Commodity_REF <- fread("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/HS10M 2022 AAFC REF.csv",
                       colClasses = c(HS10_Code = "character"))


setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")


# Loop to load and write the data

for (i in 1988:2002) {
  
  # Paste year (1988-2023) to standardized file path to create file name

  Import_Filename <- paste0("ODPFN014_", i, "12C.csv", sep="")
  
  # Load each file name for all years (1988-2002)
  
  Import <- fread(Import_Filename,
                  colClasses = c(HS10 = "character"))
  
  # Rename variables to a useable format
  
  ImportYear <- Import %>%
    rename (Date = `YearMonth/AnnéeMois`,
            HS10_Code = `HS10`,
            Country = `Country/Pays`,
            State = `State/État`,
            Value = `Value/Valeur`,
            Quantity = `Quantity/Quantité`,
            UOM = `Unit of Measure/Unité de Mesure`
            ) %>%
    
    # Merge data with Commodity_REF to filter for agriculture, seafood and food manufacturing HS Codes only
    
    merge(Commodity_REF,by="HS10_Code") %>%
    
    # Drop all other variables other than 1:8
    
    select(1:8)
  
  Import_Outfile_Start <- paste("Import", i, sep="")
  
  Import_Outfile <- paste(Import_Outfile_Start,".csv", sep="")
  
  fwrite(ImportYear, Import_Outfile)
}

toc()
```

# Append all years into 1 file

The preceding code chunk has downloaded, unzipped and extracted all files from 1988 to the most recent year. The following chunks load and bind all years:

1.  Exports

2.  Imports

3.  Trade

## 1. Exports: Load and Bind All Years

The code chunk under this heading is an example of how to append all annual ***import*** data frames found in the [Annual](%22\sfp.idir.bcgov\S104\S04004\Statistics%20and%20Research\Raw%20Data%20and%20Information\Federal%20Government%20Sources\Canadian%20International%20Merchandise%20Trade\Annual%22) folder and write a csv file file for data visualization.

```{r Bind_ExportAnnual}

tic()

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")

# Initialize an empty list to store data frames
export_list <- list()

# Loop through the years 2013 to 2023

for (year in 2003:2023) {
  # Construct the file name
  file_name <- paste0("Export", year, ".csv")
  
  # Read the CSV file
  Exportyear <- fread(file_name, colClasses = c(HS8_Code = "character"))
  
  # Add the data frame to the list
  export_list[[length(export_list) + 1]] <- Exportyear
}

# Bind all data frames in the list into one data frame
ExportFULL <- bind_rows(export_list, .id = NULL)

toc()
```

```{r Write_ExportFULL}

tic()

# Create the ImportFULL dataframe with Date, Year, and Month columns
ExportFULL <- ExportFULL %>%
  mutate(
    Year = substr(Date, 1, 4),       # Extract the first 4 characters as Year
    Month = substr(Date, nchar(Date)-1, nchar(Date))  # Extract the last 2 characters as Month
  )

ExportFULL$Date <- ymd(paste0(ExportFULL$Year,ExportFULL$Month,"01", sep = "-"))

#drop year and month columns
ExportFULL <- ExportFULL[, -c(9, 10)]

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")

fwrite(ExportFULL, "ExportFULL.csv")

toc()
```

## 2. Imports: Load and Bind All Years

The code chunk under this heading is an example of how to append all annual ***import*** data frames found in the [Annual](%22\sfp.idir.bcgov\S104\S04004\Statistics%20and%20Research\Raw%20Data%20and%20Information\Federal%20Government%20Sources\Canadian%20International%20Merchandise%20Trade\Annual%22) folder and write a csv file file for data visualization.

```{r Bind_ImportAnnual}
tic()  

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")


# Initialize an empty list to store data frames
import_list <- list()  

# Loop through the years 1988 to 2023 
for (year in 2003:2023) 
  {   
  # Construct the file name   
  file_name <- paste0("Import", year, ".csv")      
  # Read the CSV file   
  Importyear <- fread(file_name, colClasses = c(HS10_Code = "character"))      
  # Add the data frame to the list   
  import_list[[length(import_list) + 1]] <- Importyear }  

# Bind all data frames in the list into one data frame 
ImportFULL <- bind_rows(import_list, .id = NULL)  

toc()
```

```{r Write_ImportFULL}
tic() 

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")

# Create the ImportFULL dataframe with Date, Year, and Month columns 
ImportFULL <- ImportFULL %>%
  mutate(Year = substr(Date, 1, 4),  
         
# Extract the first 4 characters as Year
    Month = substr(Date, nchar(Date)-1, nchar(Date))) 

# Extract the last 2 characters as Month
    
ImportFULL$Date <- ymd(paste0(ImportFULL$Year,ImportFULL$Month,"01", sep = "-")) 

#drop year and month columns 

ImportFULL <- ImportFULL[, -c(9, 10)]  

fwrite(ImportFULL, "ImportFULL.csv")  
                                           
toc() 
```

## 3. Trade: Full Join Export FULL and Import FULL

Need to merge both and calculate trade balance

```{r Read_ExpImpFULL}

# Run if you do not have ExportFULL and ImportFULL in the environment

tic()

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")

ExportFULL <- fread("ExportFULL.csv", colClasses = c(HS8_Code = "character"))
ImportFULL <- fread("ImportFULL.csv", colClasses = c(HS10_Code = "character"))

toc()
```

```{r Clean_ImportFULL}

tic()

# Extract the first 8 characters as HS8 for Imports
ImportFULL_HS8 <- ImportFULL %>%
  mutate(
    HS8_Code = substr(HS10_Code, 1, 8))       

# Remove HS10_Code
ImportFULL_HS8 <- ImportFULL_HS8[, -c(1)]

# Group by HS8_Code
ImportFULL_HS8 <- ImportFULL_HS8 %>% 
  group_by(HS8_Code, 
           Date, 
           Country, 
           Province, 
           State, 
           UOM) %>% 
  summarise(Value = sum(Value),
            Quantity = sum(Quantity))


# Format as Date

ImportFULL_HS8$Date <- as.Date(ImportFULL_HS8$Date)
ExportFULL$Date <- as.Date(ExportFULL$Date)

toc()
```

```{r Join_TradeFULL}

tic()

TradeFULL <- full_join(ExportFULL, ImportFULL_HS8,
                       by = c("HS8_Code",
                              "Date",
                              "Country",
                              "Province",
                              "State",
                              "UOM"),
                       copy = FALSE,
                       suffix = c("_X","_M"),
                       keep = NULL)

#replace NAs with 0

TradeFULL[is.na(TradeFULL)] <- 0

# calculate trade balance

TradeFULL$Value_Balance <- TradeFULL$Value_X-TradeFULL$Value_M

TradeFULL$Quantity_Balance <- TradeFULL$Quantity_X-TradeFULL$Quantity_M

# Re-arrange the columns in a useful order
TradeFULL <- TradeFULL[,c(1:5,8,6:7,9:12)]


TradeFULL <- pivot_longer(TradeFULL,
                          7:12,
                          names_to = "Estimate",
                          values_to = "Value")

toc()
```

```{r write_TradeFULL}

tic()

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")


fwrite(TradeFULL, "TradeFULL.csv",
       row.names = FALSE,
       col.names = TRUE,
       buffMB = 10)

toc()
```
